---
title: Click here to view my disintegration as I attempt to explain R and Python
author: ''
date: '2020-12-11'
slug: click-here-to-view-my-disintegration-as-i-attempt-to-explain-r-and-python
categories: []
tags: []
description: ''
---



Hello there.

I'm going to attempt to demonstrate the extend of my coding skills below. Prepare to be blown away. (Only if there's a monstrous wind blowing and you're sitting outside. If not, the threat is completely bogus and you can ignore it.)

 
Here we have a code chunk in R, a programming language, which is essentially a way to communicate with the computer (for my non-computer science friends out there-- I was you recently, or still am?)

I've created a vector that captures the main message of this blog.
```{r}
#R code chunk
library(reticulate)

hello <- c("you", "really", "shouldn't", "have", "clicked.")
```


This is a code chunk in python, another programming language. 
```{python}
#Python code chunk
r.hello
```
*Python prints the code inputted in R.*

```{python}
#Python code chunk
however= ("Isn't this kinda marvelous?")
```


```{r}
#R code chunk
py$however
```
*R Prints the code in Python. Cool or what?*

They are communicating with each other. By golly, how?

There are smarter people than me that can explain it. I suggest you stop wasting your time and go find them.
